---
title: "Evaluating Biomarkers in Medicine: Methods and Practice"
author: "Michael Sachs"
date: March 6, 2015
output: 
  ioslides_presentation:
    self_contained: true
    widescreen: true
    fig_width: 8
    fig_height: 4
    logo: US-NIH-NCI-Logo.svg
    css: styles.css
---

# Introduction

## Evaluating Biomarkers: (Bad) Practice

```{r libs, include = FALSE}
library(ggplot2)
library(plyr)
library(ggthemes)
library(survival)
library(survivalROC)
library(plotROC)

ggkm <- function(sfit, xlabs = "Time since marker measured", ylabs = "survival probability", 
                 ystratalabs = NULL, ystrataname = NULL, timeby = 100, 
                 main = "Kaplan-Meier Plot", pval = TRUE, ...) {

if(is.null(ystratalabs)) {
   ystratalabs <- as.character(levels(summary(sfit)$strata))
}
m <- max(nchar(ystratalabs))
if(is.null(ystrataname)) ystrataname <- "Strata"
times <- seq(0, max(sfit$time), by = timeby)
.df <- data.frame(time = sfit$time, n.risk = sfit$n.risk,
    n.event = sfit$n.event, surv = sfit$surv, strata = summary(sfit, censored = T)$strata,
    upper = sfit$upper, lower = sfit$lower)
levels(.df$strata) <- ystratalabs
zeros <- data.frame(time = 0, surv = 1, strata = factor(ystratalabs, levels=levels(.df$strata)),
    upper = 1, lower = 1)
.df <- rbind.fill(zeros, .df)
d <- length(levels(.df$strata))
p <- ggplot(.df, aes(time, surv, group = strata)) +
    geom_step(aes(linetype = strata), size = 1) +
    theme(legend.position = c(.65, .5), legend.key = element_rect(colour = NA)) +
    labs(linetype = ystrataname) + 
    scale_y_continuous(ylabs, limits = c(0, 1)) +
    labs(linetype = ystrataname) 
 
## Plotting the graphs
p
   
}
```



```{r plot3, echo = FALSE, fig.width = 4.5, fig.height = 4.5, warning = FALSE}
D <- rbinom(250, 1, .3)
T <- rnorm(250, mean = D, sd = .75)
  
c <- sort(T)
TPF <- sapply(c, function(x) mean(T[D == 1] > x))
FPF <- sapply(c, function(x) mean(T[D == 0] > x))

badroc <- data.frame(cbind(c, TPF, FPF))

boxplot(T ~ D, xlab = "disease status", ylab = "marker value")
text(1.5, -1.5, "p < .001")
plot(TPF ~ FPF, badroc, lwd = 2, xlab = "1 - specificity", ylab = "sensitivity", type = 's', xaxp = c(0, 1, 2), yaxp = c(0, 1, 2))
legend(.55, .2, lwd = 2, lty = 1, legend = "Great new marker\n AUC = .79; p < 0.001 \n", cex = .75)
```

## Evaluating Biomarkers: (Bad) Practice

```{r plot2, echo = FALSE, fig.width = 6, fig.align = 'center', warning = FALSE}
df <- data.frame(Time = c(rexp(50, .1), rexp(50, 2), rexp(50, 4), rexp(50, 3)), 
                 Biomarker = c(rep("Below median", 100), rep("Above median", 100)), 
                 Treatment = rep(c(rep("Experimental", 50), rep("Placebo", 50)), 2))


df <- ddply(df, ~ Biomarker, function(d){
  
  data.frame(Time = d$Time, Survival = (nrow(d):1/nrow(d))[rank(d$Time, ties.method = "first")])
  
})

ggplot(df, aes(x = Time, y = Survival, color = Biomarker)) + scale_x_continuous(limits = c(0, 10)) +
  geom_step(lwd = 1.5) + theme_igray() + annotate("text", 7.5, .1, label = "* p < 0.001")

```


## Evaluating Biomarkers: (Bad) Practice

```{r plot1, echo = FALSE, fig.align='center', warning=FALSE}

df <- data.frame(Time = c(rexp(50, .1), rexp(50, 1), rexp(50, .9), rexp(50, 1.5)), 
                 Biomarker = c(rep("Below median", 100), rep("Above median", 100)), 
                 Treatment = rep(c(rep("Experimental", 50), rep("Placebo", 50)), 2))

df <- ddply(df, ~ Biomarker + Treatment, function(d){
  
  data.frame(Time = d$Time, Survival = (nrow(d):1/nrow(d))[rank(d$Time, ties.method = "first")])
  
})

ann.text <- data.frame(Time = 4, Survival = 1, Treatment = "Experimental", Biomarker = c("Below median", "Above median"), label = c("p < 0.001", "p = 0.28"))

ggplot(df, aes(x = Time, y = Survival, color = Treatment)) + scale_x_continuous(limits = c(0, 5)) +
  geom_step() + theme_igray() + scale_color_colorblind() + 
  geom_text(data = ann.text, aes(label = label), color = "black") + facet_wrap(~Biomarker)
```

## Evaluating Biomarkers: Theory


Search Term | Results
-----------|-------------
Clinical Trial(s) | `r 2140 + 602 + 931 + 327 + 478 + 875 + 1210 + 3030`
Survival | `r 2650 + 1910 + 1260 + 533`
Longitudinal | `r 306 + 876 + 1250 + 1900`
**Biomarker(s)** | **`r 333 + 394 + 216 + 174 + 206 + 250 + 71 + 53`**
Genomic(s) | `r 61 + 42 + 235 + 340 + 233 + 170 + 141 + 182`


Search results from Google scholar restricted to journals: _Statistics in Medicine, Biometrics, Biostatistics, Biometrical Journal_


## About me 

- Mathematical Statistician at the NCI
    + Review biomarker-related aspects of cancer clinical trials
    + Collaborate with other programs across the NCI and NIH
    + Conduct research in Biostatistics

Disease areas: 

- Prostate, AIDS-related, Head & Neck, Bladder/Renal/Colon, Melanoma, Pancreatic, CNS/Brain
- Previous experience in kidney disease, diabetes, heart disease, Alzheimer's disease

Other interests:

- Reproducible research, statistical computing, data visualization
  
# The Basics

## Uses of biomarkers/imaging tests in oncology

  - Screening (mammography, HPV test)
  - Diagnosis (HER-2, PSA)
  - Prognosis, predicting future recurrence (Oncotype, Mammaprint)
  - Treatment selection (ER, ALK, KRAS)
  - Surrogate endpoints (tumor growth rate, MRD)
  
[Source (cancer.gov)](http://www.cancer.gov/cancertopics/factsheet/detection/tumor-markers)  

## Biomarker statistics 101

Binary biomarker and Binary outcome

<style>
td { border: 1px #aaa solid;
     padding: .2em; 
     }
</style>

<table align="center" style="text-align:center; border:1px; background:transparent; font-size: medium;">
<tr>
<td colspan="2" style="border:none;"></td>
<td colspan="2" style="background:#eeeebb;"><b>Disease or Event</b></td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="background:#ddddcc;">Total population</td>
<td style="background:#ffffcc;">Disease positive</td>
<td style="background:#ddddaa;">Disease negative</td>
<td style="background:#ddddcc;">Prevalence =<br />
Σ&#160;Condition positive
<div style="border-top:1px solid;">Σ&#160;Total population</div>
</td>
</tr>
<tr>
<td rowspan="2" style="background:#bbeeee;"><b>Test<br />
outcome</b></td>
<td style="background:#ccffff;">Test<br />
outcome<br />
positive</td>
<td style="background:#ccffcc;"><span style="color:#006600;"><b>True positive</b></span></td>
<td style="background:#eedddd;"><span style="color:#cc0000;"><b>False positive</b></span><br />
</td>
<td style="background:#ccffff;"> Positive predictive value (PPV) =<br />
Σ True positive
<div style="border-top:1px solid;">Σ&#160;Test&#160;outcome&#160;positive</div>
</td>
</tr>
<tr>
<td style="background:#aadddd;">Test<br />
outcome<br />
negative</td>
<td style="background:#eedddd;"><span style="color:#cc0000;"><b>False negative</b></span><br />
</td>
<td style="background:#bbeebb;"><span style="color:#006600;"><b>True negative</b></span></td>
<td style="background:#aadddd;">Negative predictive value (NPV) =<br />
Σ True negative
<div style="border-top:1px solid;">Σ&#160;Test&#160;outcome&#160;negative</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="border:none;"></td>
<td style="background:#ffffcc;">True positive fraction (TPF, Sensitivity) =<br />
Σ True positive
<div style="border-top:1px solid;">Σ&#160;Disease&#160;positive</div>
</td>
<td style="background:#ddddaa;">False positive fraction (FPF) =<br />
Σ False positive
<div style="border-top:1px solid;">Σ&#160;Disease&#160;negative</div>
</td>
<td style="background:#ddddcc;">Accuracy (ACC) =<br />
Σ&#160;True positive + Σ True negative
<div style="border-top:1px solid;">Σ&#160;Total population</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
<td style="border:none;"></td>
<td style="background:#ffffcc;">False negative fraction (FNF) =<br />
Σ False negative
<div style="border-top:1px solid;">Σ&#160;Disease&#160;positive</div>
</td>
<td style="background:#ddddaa;">True negative fraction</a> (TNF, Specificity) =<br />
Σ True negative
<div style="border-top:1px solid;">Σ&#160;Disease&#160;negative</div>
</td>
</tr>
<tr>
<td style="border:none;"></td>
</tr>
</table>

## Probabilities and Proportions

**Quality of test-based classifications:**

$$ TPF = Pr\{\mbox{Marker positive} | \mbox{Has condition} \} = Pr\{M = 1 | D = 1\} $$

$$ FPF = Pr\{M = 1 | D = 0 \} $$ 

Ideal: TPF = 1, FPF = 0; Bad: TPF = FPF

**Predictions**

$$ PPV = Pr\{D = 1 | M = 1\} $$

$$ NPV = Pr\{D = 0 | M = 0\} $$
 
Ideal: PPV = NPV = 1; Bad: PPV = $Pr\{D = 1\}$; NPV = $Pr\{D = 0\}$
 
## One statistic not enough

**All of the information in that table can be obtained given only 3 of the quantities: **

 - Usually, $(TPF, FPF, Pr\{D = 1\})$
 - Occassionally, $(PPV, NPV, Pr\{M = 1\})$

Why? Bayes Rule: $Pr\{M = 1 | D = 1\} =$

$$ \frac{Pr\{M = 1 \mbox{ AND } D = 1\}}{Pr\{D = 1\}} = \frac{Pr\{D = 1 | M = 1\} Pr\{M = 1\}}{Pr\{D = 1\}} $$
 
 - Most commonly, only a HR or OR is reported
 - HR/OR are insufficient to evaluate a test proposed for use in individuals ([Pepe et al. 2004 AJE](http://aje.oxfordjournals.org/content/159/9/882.short))

# ROC Curves

## Extension 1

What if the test yields a continuous measurement?

```{r mamm, echo = FALSE, warning = FALSE, message = FALSE}
onc <- read.csv("nki-oncotype-data-2015-02-27.csv", stringsAsFactors = FALSE)

onc$o4 <- onc$oncotypedx
onc$o4[onc$oncotypedx == 100] <- onc$oncotypedx[onc$oncotypedx == 100] - rexp(sum(onc$oncotypedx == 100), 3.5)
onc$o4[onc$oncotypedx == 0] <- onc$oncotypedx[onc$oncotypedx == 0] + rexp(sum(onc$oncotypedx == 0), 3.5)

onc$oncotypedx <- onc$o4

onc$d.5year <- with(onc, t.rfs < 365.25 * 5 & e.rfs == 1)
onc$d.5year[with(onc, t.rfs < 365.25 * 5 & e.rfs == 0)] <- NA
onc$riskmedian <- onc$oncotypedx > median(onc$oncotypedx)

contex <- data.frame(Result = c(onc$oncotypedx, rbinom(250, 1, .15)), Test = c(rep("Oncotype DX score", 329), rep("Mammogram", 250)))
ggplot(contex, aes(x = Result)) + geom_bar(stat = "bin") + facet_wrap(~ Test, scales = "free") + theme_igray() + scale_color_colorblind()
```

## Oncotype DX

- Risk score for recurrence of breast cancer
- Based on expression of a panel of 21 genes
- Measured using RT-PCR on archival tumor specimens
- Example data (n = 207) from a clinical case series in the Netherlands
- Outcome is recurrence within 5 years

```{r oncsurv, echo = FALSE, fig.width = 4.25, fig.height = 2.5}
par(mar = c(4, 4, 2, 1), cex = .75)
plot(survfit(Surv(t.rfs, e.rfs) ~ 1, data = onc), xscale = 365.25, mark.time = FALSE, conf.int = FALSE, lwd = 2, xlab = "Years post-surgery", ylab = "RFS")
abline(v = c(2, 5, 10), h = seq(0, 1, by = .1), col = "grey80")

plot(survfit(Surv(t.os, e.os) ~ 1, data = onc), xscale = 365.25, mark.time = FALSE, conf.int = FALSE, lwd = 2, xlab = "Years post-surgery", ylab = "OS")
abline(v = c(2, 5, 10), h = seq(0, 1, by = .2), col = "grey80")
```

References: [Paik et al., 2004](http://www.nejm.org/doi/full/10.1056/NEJMoa041588), [van de Vijver et al., 2002](http://www.nejm.org/doi/full/10.1056/nejmoa021967), [van't Veer et al., 2002](http://www.nature.com/nature/journal/v415/n6871/full/415530a.html)

## Solutions

### Goal is to assess accuracy using the aforementioned measures

> - Pick an arbitrary cutoff, say $c$. 
> - Now *test positive* is defined as $M > c$
> - Consider measures as functions of $c$, i.e.
    $$ TPF(c) = Pr\{M > c | D = 1\} $$
    $$ FPF(c) = Pr\{M > c | D = 0\} $$
    
##

<iframe src="https://sachsmc.shinyapps.io/illustrateROC/"></iframe>

## ROC Curve Plots

Enable the observer to 

  - Accurately discern the operating charactersitics at any cutoff
  - Guide the choice of "optimal" cutoff
  - Determine what the cutoff is! 

A short review of the oncology literature suggests

  - Most ROC curve plots do not acheive these aims
  - probably due to limitations of plotting software
  
## My solution

```{r rocci, echo = 1, results = 'asis', fig.keep = 'none', fig.align = 'center'}
library(plotROC)
onc.sub <- subset(onc, !is.na(d.5year))

D.ex <- onc.sub$d.5year
M.ex <- onc.sub$oncotypedx

roc.ci <- calculate_roc(M.ex, D.ex, ci = TRUE, alpha = 0.05)
ci.rocplot <- ggroc(roc.ci, label = "Oncotype DX", ci = TRUE, label.adj.x = .05)
cat(
  export_interactive_roc(ci.rocplot + theme_igray(), 
                         prefix = "aci")
  )
```

## Print use

See http://sachsmc.github.io/plotROC for usage and examples

```{r printci, fig.align = "center", echo = FALSE, fig.height = 5, fig.width = 5}
plot_journal_roc(ci.rocplot, n.cuts = 10, 
                 ci.at = c(25, 50, 75))
```

## Extension 2

  - Binary outcome, good for diagnosis, screening
  - What about prognostic/predictive biomarkers?
  - Outcome of interest is *time to event*, i.e. PFS or OS
  - Instead of $D = 1$ we have $T > t$
  
```{r surv, fig.height = 3, fig.width = 5, echo = FALSE, fig.align = 'center'}
ggplot(data.frame(T = sort(exp(rnorm(100, mean = 1, sd = .2)))), aes(x = T, y = 1 - seq_along(T)/100)) + geom_step() + 
  scale_x_continuous("Time since surgery") +
  scale_y_continuous("Proportion without distant recurrence") + theme_igray()
```
  
## Solutions

> - Dichotomize time and proceed as binary
>     + possibly throwing out data (censoring)
> - Use the Kaplan-Meier estimate 
>     + Time-dependent TPF and FPF
      $$ TPF(c, t) = P(M > c | T \leq t) $$
      $$ FPF(c, t) = P(M > c | T > t) $$

## Time dependent TPF and FPF

A little algebra (Bayes rule again): 

$$ TPF(c, t) = \frac{P(T \leq t | M > c) P(M > c)}{P(T \leq t)} $$

The KM curve is an estimate of $P(T > t)$, and $P(T > t | M > c)$ in the subgroup with $M > c$. 

Combine that with the proportion of $M > c$ and we are done. 

See [Heagerty, Lumley, Pepe, 2000](http://onlinelibrary.wiley.com/doi/10.1111/j.0006-341X.2000.00337.x/abstract) for more details. 

## The plot
Show curves for a small subset of event times:
<div class="columns-2">
```{r survget, echo = FALSE, warning = FALSE, message = FALSE, results = 'asis', fig.keep = 'none', fig.width = 5, fig.height = 5}
library(survivalROC)
library(survival)
set.seed(320)

onc.sub2 <- subset(onc, !is.na(t.rfs))
survT <- onc.sub2$t.rfs/365.25
cens <- onc.sub2$e.rfs

M <- onc.sub2$oncotypedx

oncotypeDX <- factor(M < median(M), labels = c("above median", "below median"))

sfit <- survfit(Surv(survT, cens) ~ oncotypeDX)
```

```{r surv2, echo = FALSE, warning = FALSE, message = FALSE, fig.width = 4.5, fig.height = 4.5}
ggkm(sfit) + scale_x_continuous("Time since surgery") + theme_igray() + 
  theme(legend.position = c(.75, .25)) + scale_y_continuous("Proportion without distant recurrence", limits = c(0, 1))
```


```{r survplot, echo = FALSE, warning = FALSE, message = FALSE, results = 'asis', fig.keep = 'none', fig.width = 5, fig.height = 4}
sroc <- lapply(c(2, 5, 10), function(t){ 
  stroc <- survivalROC(Stime = survT, status = cens, marker = M, predict.time = t, method = "NNE", span = .25 * 350^(-.2))
  data.frame(TPF = stroc$TP, FPF = stroc$FP, c = stroc$cut.values, time = rep(stroc$predict.time, length(stroc$FP)))
  })


names(sroc) <- c("2 years", "5 years", "10 years")
colorplot <- multi_ggroc(sroc, 
                         xlabel = "1 - Specificity", 
                         ylabel = "Sensitivity",
                         label = names(sroc), 
                         label.adj.x = c(-.2, -.1, 0), 
                         label.adj.y = c(0, 0, 0), legend = TRUE,
                         ) + theme_igray() + geom_abline(slope = 1, intercept = 0, color = "grey80")



cat(
  export_interactive_roc(colorplot, lty = rep(1, 3), prefix = "multi3",
                         color = c("black", "purple", "orange"), 
                         legend = TRUE)
  )
```
</div>

## More extensions and questions

- Marker measured longitudinally ($M(t)$)
- Condition ($D$) not directly observable
- Categorical disease state (_e.g._ Cancer stage)
- Confounding/effect modification

# Prospective Accuracy

## Retrospective vs Prospective

- Retrospective: Pr(Marker value | disease status)
    + TPR, FPR, ROC Curve
    + Inherent capacity of marker to classify
    + Does not depend on disease prevalence
- Prospective: Pr(Disease or Event | Marker Value)
    + PPV, NPV, condition on marker threholds
    + **Predictiveness curve** conditions on marker value
    + Risk prediction and medical decision making
    + Depend on event rate/disease prevalence
    
    
## Predictiveness curve

> $$ R(v) = Pr\{D = 1 | F(M) = v\}, $$ where $F$ is the CDF of $M$. 

Equivalent to the quantile function of $Pr\{D = 1 | M\}$. To estimate: 

- Model $Pr\{D = 1 | M\}$ as $G(\beta; M)$ in some way, 
- Estimate $F$ empirically, $\hat{F}(m) = n^{-1}\sum 1[M_i < m]$
- Then $\hat{R}(v) = G\{\hat{\beta}; \hat{F}^{-1}(M)\}$.


##

<iframe src="https://sachsmc.shinyapps.io/predcurver/"></iframe>

## Summary measures

> Total gain (TG): 
> $$ TG = \int_{0}^1 | R(t) - \theta | \; dt, $$
> where $\theta = Pr\{D = 1\}$. 

Area between the curve and the horizontal line

> Proportion of explained variation (PEV):
> $$ PEV = \frac{1}{\theta (1 - \theta)} \int_{0}^1 (p - \theta)^2 \; dR^{-1}(p). $$

Amount of spread in the vertical axis

## Partial summary measures

- Quantify the steepness of the predictiveness curve
- Allow for hypothesis testing, comparison of biomarkers

In the presence of low risk/high risk thresholds, calculate partial summary statistics, i.e.

```{r psum, fig.width = 4, fig.height = 4, fig.align = 'center', echo = FALSE}
library(predcurver)
M.ex <- onc.sub$oncotypedx
D.ex <- onc.sub$d.5year

rho <- predict(glm(D.ex ~ M.ex, family = binomial), type = "response")
pr <- predcurve(rho)

xxx <- pr$percentile[pr$percentile < .25]
xxx <- c(xxx, xxx[length(xxx)], rev(xxx))

yyy <- pr$risk[pr$percentile < .25]
yyy <- c(yyy, rep(mean(D.ex), length(yyy) + 1))

xxx2 <- pr$percentile[pr$percentile > .75]
xxx2 <- c(xxx2, xxx2[length(xxx2)], rev(xxx2))

yyy2 <- pr$risk[pr$percentile > .75]
yyy2 <- c(yyy2, rep(mean(D.ex), length(yyy2) + 1))


ggplot(pr, aes(x = percentile, y = risk)) + geom_line() + 
  scale_y_continuous("Risk 5 year recurrence", limits = c(0, 1)) + 
  scale_x_continuous("Oncotype DX percentile") + 
  geom_hline(yintercept = mean(D.ex), linetype = 2) + theme_igray() + 
  geom_polygon(data = data.frame(xxx, yyy), aes(x = xxx, y = yyy), alpha = .5) + 
  geom_polygon(data = data.frame(xxx2, yyy2), aes(x = xxx2, y = yyy2), alpha = .5)
```

## Predictiveness curve summary

- Asymptotic distributions of summary measured calculated using functional delta method
- Permutation-based test statistics to comparing biomarkers

```{r eval = FALSE}
devtools::install_github("sachsmc/predcurver")
library(predcurver)
```

Easy-to-use `R` class for estimation, testing, summarizing, and plotting. 

See [Sachs and Zhou, 2013](http://onlinelibrary.wiley.com/doi/10.1002/bimj.201200146/) for more details

[Janes et al. 2014](http://www.ncbi.nlm.nih.gov/pubmed/24695044) for extentions to biomarkers for treatment selection

# Surrogate endpoints

## Definition

Thus far, we've considered biomarkers measured pre-treatment, or in the absence of treatment. 

Purpose of a surrogate biomarker is to predict the likely treatment effect on a clinical outcome: 

```{r pic, echo = FALSE}
library(DiagrammeR)
diagram <- "
graph LR
  A(Treatment    .)-->B(Surrogate   .)
  A(Treatment    .)-->C(Outcome    .)
  B(Surrogate    .)-->C(Outcome    .)

"
mermaid(diagram)
```

## Example: Herpes Zoster

- Phase III randomized trial of Zoster vaccine vs placebo in 22,439 North American/European people aged 50-59
- Vaccine highly effective: VE = 70% (95% CI: 54% to 81%)
- Antibodies to vaccine (titers) measured in blood post-randomization in a case-control substudy
- Possible correlates of protection: 
    + Titers at week 6
    + Fold change in Titers from baseline
- Vaccine efficacy as function of biomarker
    
Reference: [Gilbert, Gabriel, et al, 2014](http://jid.oxfordjournals.org/content/210/10/1573.short)


## Modeling approaches

- As before, interested in prospective measure, but this time **treatment-specific**: 

$$ Pr\{T < t(0) | M = m(1)\} - Pr\{T < t(0) | M = m(1)\}, $$

$=$ the conditional risk difference, where $t(0)$ is the event time under placebo and $t(1)$ is event time under active treatment 

- Some assumptions and special methods allow for estimation of causal effects
- For a given subject, can only observe either $t(0)$ or $t(1)$, never both
- Principal stratification, counterfactuals



## Summary curves

```{r riskdiff, echo = FALSE, fig.width = 5, fig.height = 5, fig.align = 'center'}
M.ex <- rnorm(250)
M2.ex <- rnorm(250)
D.ex <- rbinom(250, 1, pnorm(-2 + 3 * M.ex + .25 * M2.ex))

rho <- predict(glm(D.ex ~ M.ex, family = binomial), type = "response")
rho2 <- predict(glm(D.ex ~ M2.ex, family = binomial), type = "response")
pr <- predcurve(rho)
pr2 <- predcurve(rho2)

pr$Surrogate <- "Log difference in Titers"
pr2$Surrogate <- "Titers at week 6"

ggplot(rbind(pr, pr2), aes(x = percentile, y = risk, linetype = Surrogate)) + geom_line() + 
  geom_hline(yintercept = mean(D.ex), linetype = 2) + theme_igray() + 
  theme(legend.position=c(.25, .85)) + scale_y_continuous("Risk difference (vaccine efficacy)") + 
  scale_x_continuous("Surrogate percentile on treatment")
```

## Details

- [Gabriel et al. 2015](http://onlinelibrary.wiley.com/doi/10.1002/sim.6349/) use a novel Weibull model to estimate causal risks in the time-to-event setting. 
    + Transformations to get retrospective summaries, $TPF$ and $FPF$. 
    + Also summary statistics such as $TG$ and $PEV$ to compare surrogates
- Flexible model that allows for time variation in hazard and treatment effect

```{r waning, fig.width = 6, fig.height = 3, echo = FALSE, fig.align = 'center'}
time <- sort(runif(250 ,0, 10))
tg <- c(.65 - .05 * time + .025 * (time - 7) * (time > 7), .25 - .01 * time)

ggplot(data.frame(time = c(time, time), tg = tg, Surrogate = c(rep("Log difference in Titers", 250), rep("Titers at week 6", 250))), 
       aes(x = time, y = tg, linetype = Surrogate)) + 
  geom_line() + scale_y_continuous("Total Gain", limits = c(0, 1)) + 
  scale_x_continuous("Time since surrogate measurement") + theme_igray()
```

# Conclusion

## 

- Methods and tools to comprehensively evaluate biomarkers of all types
- Frame and visualize the data and statistics that researchers care about
- Arbitrary cutoffs and p-value thresholds hinder accumulation of knowledge and the interpretation of biomarkers
- Embrace new technologies to make these tools available to the scientific community

### **Thank you!**

Presentation and source files available at 

http://github.com/sachsmc/biomarkers-in-medicine


